Meta-analytic approaches and 'new' effect sizes to account for heterogeneous effects in comparative physiology

### 2021-07-07 MS/ms.Rmd
Making some revisions

A meta-analysis, being part of systematic review, provides quantitatively informed answers to research questions by synthesising research results across studies and identify sources of variation [@Nakagawa2017; @Gurevitch1999; @Gurevitch2018; @Cooper2009; @Arnqvist1995; @Koricheva2013]. Meta-analysing independent studies provides greater statistical power and precision than what any individual study on its own would be able to provide on a question - particularly given that most empirical studies are already under-powered in many areas of biology [@Forstmeier2017; @Jennions2003; @Button2013]. By expressing study effects on a common scale (i.e., standardised effect size) a meta-analyst can gain broader insight into the direction and efficacy of a particular treatment effect or the strength of a relationship between two variables of interest [@Gurevitch2018; @Koricheva2013]. Quantitative syntheses, including meta-analyses, have provided comparative physiologists with powerful insights on pressing global challenges - from testing whether physiological plasticity can buffer organisms against climate change [e.g, @Seebacher2015] to the degree to which endocrine disrupting chemicals, such as bis phenol A (BPA), impact aquatic organisms [e.g., @WuSeebacher2020].

Meta-analysis is a well-developed research topic with a long history in statistics, medicine, psychology and ecology [@Gurevitch2018; @Nakagawa2017; @Glass2015]. It has emerged as the 'gold-standard' across various disciplines for quantitative synthesis and is becoming increasingly more prominent in the field of comparative physiology. Formal meta-analyses, as a part of a systematic review, involve a systematic literature-based search, extraction of data from studies meeting specific inclusion criteria, and the subsequent calculation of effects sizes from summary statistics provided by papers. The entire process should follow and report upon this entire process carefully to ensure the meta-analysis within a systematic review, is as transparent and reproducible as possible [i.e., see the Preferred Reporting Items for Systematic Reviews and Meta-Analyses in Ecology and Evolution (PRISMA-EcoEvo) protocol, @ODea2021]. Aside from providing an up-to-date overview of the research question, along with identifying any gaps and biases, there are two major goals of meta-analysis: 1) to provide an overall mean estimate of a treatment effect or relationship and 2) to quantify effect size variance and understand key drivers explaining differences in effects across studies [@Nakagawa2017; @Gurevitch1999; @Gurevitch2018; @Cooper2009; @Koricheva2013]. A critically important feature of meta-analysis is its ability to account for sampling error [@Koricheva2013]. The inverse of sampling error variance is used to weight studies, accordingly, in the analysis and make overall estimates more precise. While many meta-analyses might employ unweighted analyses, these studies will not gain greater precision on estimates because sampling variance cannot be formally accounted for in the analysis, even if the overall mean itself will be unbiased [@Morrissey2016a]. <!--- cite Morrissey here - also, the overall mean is unbiased given there is no publication bias or small study effects - this is important as many meta-analyses are likley to suffer from small study effects; my pick sann does not seem like you have talked about publcaiton bias too much - an ability of meta-analysis to detect publication bias is one of its strength - or probably the major strength...  --->

Despite the widespread adoption and well established methodological procedures of meta-analysis, it is often criticised for mixing 'apples and oranges' -- in other words, mixing effects from studies that lack comparability [@Gurevitch2018; @Arnqvist1995; @Gallo1979; @Carpenter2020; @Stewart2010; @Glass2015]. Lack of comparability could result from studies differing in experimental design, species, measurement variables and sampling units, which can impact the generalisability of meta-analytic results [@Arnqvist1995; @Stewart2010]. Indeed, study specific differences in comparative physiology are widespread. For example, experimental designs can vary in the temperatures that they apply or in the dosages of chemical modifiers used in experimental treatments [e.g., @WuSeebacher2020]. In addition, they often include a diversity of study species which can vary in responses. To the uninitiated, the concern might not be unfounded given that meta-analyses in ecology and evolution often mix highly heterogeneous effect size estimates, resulting in high variability [@Senior2016]. <!--- this last sentence does not seem to fit well or not completely logical - did you want to say -heterogeneity is normal. Actually Jessica always say in ecology meta-analysis is probably not that useful; meta-regression is what we are most interested - I think yo can make a similar point or you are already making one but make it clearer  - the first sentence is a good one but rather than focusing on "controlling" for hetero but finding out what causes hetero should our (comparative physiologists') main aim for meta-analysis. I know you are probably intending to say something like this but not coming through at the moment. Also, I note that "review generated evidence" only comes from meta-regression not meta-analysis = intercept models - as study general evidence can be equivalent to meta-analytic results --->

@@

Meta-analysis is always concerned with effect heterogeneity. In other words, the factors that drive differences in the direction and magnitude an of effect within and across studies [@Nakagawa2017; @Gurevitch2018]. The concept of heterogeneity is vitally important because it tells us how general a specific conclusion is likely to be, how much of the variance we see is the result of real biology (and not just sampling variance), and provides a way to understand what the key biological factors are leading to an effect. There are a plethora of phenomena driving effect heterogeneity, some of which are interesting and others that are less so. Generally, contributors to effect size heterogeneity fall within a number of broad categories. First, sampling variance contributes to heterogeneity; the variability in an effect because of the inherent variability in sampling a given population. We cannot sample the entire population and nor do we necessarily want to. As a consequence, we need to rely on a sub-sample to estimate a population parameter (e.g., correlation between two traits). How many independent samples we collect impacts our precision in estimating this 'true' effect. In meta-analysis we formally include sampling variance to remove this source of heterogeneity directly - hence why having sampling variances for an effect is so important. Second, there are sociological factors impacting heterogeneity. These include attitudes towards publishing non-significant results. Low powered studies that do not find an 'effect' or that find an effect in the opposite direction often lead to what is called publication bias. Third, there is real biology driving differences in effects. This is particularly true in comparative physiology where different species with varying habitats, life-histories, mating systems and reproductive modes are synthesised. As comparative physiologists, this is the stuff that gets us up every morning! Much variation could result from shared evolutionary history (i.e., phylogeny) or because of the different thermal environments species (or populations) occupy. We can model these biological factors to test predictions on what we hypothesize are key players impacting an effect improving our understanding of the biological world around us. Finally, methodological factors are also big contributors to effect variability. In some cases, these methodological factors are of direct interest. For example, the methods for measuring an outcome variable can result in different effects, and it's important we know and understand this as it can determine how experiments are designed in the future. In other cases, methodological factors might be considered 'nuisance heterogeneity' - factors we know vary, but are rather unsurprising to us. These can include differences in temperature, pH, dosages, water potential etc. that are obviously going to vary and have an impact on the magnitude of an effect. In some cases, these factors might actually not be 'nuisance' variables but rather the main interest. However, in many cases the can result in a more complicated interpretation of an effect measure then what is desired.

>>>>

A meta-analysis, being part of systematic review, provides quantitatively informed answers to research questions by synthesising research results across studies and identify sources of variation [@Nakagawa2017; @Gurevitch1999; @Gurevitch2018; @Cooper2009; @Arnqvist1995; @Koricheva2013; @Borenstein2019]. Meta-analysing independent studies provides greater statistical power and precision than what any individual study on its own would be able to provide on a question - particularly given that most empirical studies are already under-powered in many areas of biology [@Forstmeier2017; @Jennions2003; @Button2013]. By expressing study effects on a common scale (i.e., standardised effect size) a meta-analyst can gain broader insight into the direction and efficacy of a particular treatment effect or the strength of a relationship between two variables of interest [@Gurevitch2018; @Koricheva2013]. Quantitative syntheses, including meta-analyses, have provided comparative physiologists with powerful insights on pressing global challenges - from testing whether physiological plasticity can buffer organisms against climate change [e.g, @Seebacher2015] to the degree to which endocrine disrupting chemicals, such as bis phenol A (BPA), impact aquatic organisms [e.g., @WuSeebacher2020].

Meta-analysis is a well-developed research topic with a long history in statistics, medicine, psychology and ecology [@Gurevitch2018; @Nakagawa2017; @Glass2015]. It has emerged as the 'gold-standard' across various disciplines for quantitative synthesis and is becoming increasingly more prominent in the field of comparative physiology. Formal meta-analyses, as a part of a systematic review, involve a systematic literature-based search, extraction of data from studies meeting specific inclusion criteria, and the subsequent calculation of effects sizes from summary statistics provided by papers. The entire process should follow and report upon this entire process carefully to ensure the review process and data collation, are as transparent and reproducible as possible [i.e., see the Preferred Reporting Items for Systematic Reviews and Meta-Analyses in Ecology and Evolution (PRISMA-EcoEvo) protocol, @ODea2021]. Aside from providing an up-to-date overview of the research question there are three major goals of meta-analysis: 1) to provide an overall mean estimate of a treatment effect or relationship, 2) to quantify effect size variance and understand key drivers explaining differences in effects across studies and 3) attempt to identify gaps and publication biases [@Nakagawa2017; @Gurevitch1999; @Gurevitch2018; @Cooper2009; @Koricheva2013; @Borenstein2019]. A critically important feature of a meta-analysis is its ability to account for sampling error [@Koricheva2013]. The inverse of sampling error variance is used to weight studies, accordingly, in the analysis and make overall estimates more precise. While many meta-analyses might employ unweighted analyses, these studies will not gain greater precision on estimates because sampling variance cannot be formally accounted for in the analysis, even if the overall mean itself will be unbiased [@Morrissey2016a].

Despite its widespread adoption and well established methodological procedures, meta-analysis is often criticised for mixing 'apples and oranges' -- in other words, mixing effects from studies that lack comparability [@Gurevitch2018; @Arnqvist1995; @Gallo1979; @Carpenter2020; @Stewart2010; @Glass2015]. Lack of comparability could result from studies differing in experimental design, species, measurement variables and sampling units, which can impact the generalisability of meta-analytic results [@Arnqvist1995; @Stewart2010]. Indeed, study specific differences in comparative physiology are widespread. For example, experimental designs can vary in the temperatures that they apply or in the dosages of chemical modifiers used in experimental treatments [e.g., @WuSeebacher2020]. To the uninitiated, this concern might not be unfounded given that heterogenity in effects is often high in ecological and evolutionary meta-analyses [@Senior2016], telling us that the effect size varies a great deal from one population to another. However, to many, the very fact that highly heterogeneous effects exist is the core strength of meta-analysis when considered carefully [@Cooperetal2019; @Borenstein2019; @Gurevitch2018; @Glass2015]. Indeed, exploring the reasons for why effects are likely to vary across studies "...is the spice of life" [p.g., 519; @Cooperetal2019].

@@

Meta-analysis is always concerned with effect heterogeneity. In other words, the factors that drive differences in the direction and magnitude an of effect within and across studies [@Nakagawa2017; @Gurevitch2018; @Cooperetal2019; @Borenstein2019]. The concept of heterogeneity is vitally important because it tells us how general a specific conclusion is likely to be, how much of the variance we see is the result of real biology (and not just sampling variance), and provides a way to understand what the key biological factors are leading to an effect. There are a plethora of phenomena driving effect heterogeneity, some of which are interesting and others that are less so. Generally, contributors to effect size heterogeneity fall within a number of broad categories. First, sampling variance contributes to heterogeneity; the variability in an effect because of the inherent variability in sampling a given population. We cannot sample the entire population and nor do we necessarily want to. As a consequence, we need to rely on a sub-sample to estimate a population parameter (e.g., correlation between two traits). How many independent samples we collect impacts our precision in estimating this 'true' effect. In meta-analysis we formally include sampling variance to remove this source of heterogeneity directly - hence why having sampling variances for an effect is so important. Second, there are sociological factors impacting heterogeneity. These include attitudes towards publishing non-significant results. Low powered studies that do not find an 'effect' or that find an effect in the opposite direction often lead to what is called publication bias. Third, there is real biology driving differences in effects. This is particularly true in comparative physiology where different species with varying habitats, life-histories, mating systems and reproductive modes are synthesised. As comparative physiologists, this is the stuff that gets us up every morning! Much variation could result from shared evolutionary history (i.e., phylogeny) or because of the different thermal environments species (or populations) occupy. We can model these biological factors to test predictions on what we hypothesize are key players impacting an effect improving our understanding of the biological world around us. Finally, methodological factors are also big contributors to effect variability. In some cases, these methodological factors are of direct interest. For example, the methods for measuring an outcome variable can result in different effects, and it's important we know and understand this as it can determine how experiments are designed in the future. In other cases, methodological factors might be considered 'nuisance heterogeneity' - factors we know vary, but are rather unsurprising to us. These can include differences in temperature, pH, dosages, water potential etc. that are obviously going to vary and have an impact on the magnitude of an effect. In some cases, these factors might actually not be 'nuisance' variables but rather the main interest. However, in many cases the can result in a more complicated interpretation of an effect measure then what is desired.
